Smoothing seismicity techniques applied to seismic sources characterization and probabilistic seismic hazard analysis in Brazil

### 2015-01-13 manuscript.tex
adding images

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% American Geophysical Union (AGU)
% LaTeX Template
% Version 1.0 (3/6/13)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% The AGUTeX class and agu-ps referencing style were created and are owned 
% by AGU: http://publications.agu.org/author-resource-center/author-guide/latex-formatting-toolkit/
%
% This template has been modified from the blank AGU template to include
% examples of how to insert content and drastically change commenting. The
% structural integrity is maintained as in the original blank template.
%
% Important notes: 
% This template retains extensive commenting from the AGU template. It is heavily 
% advised you read these comments and follow them in order to insure a speedy 
% submission process.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% See how to enter figures and tables at the end of the article, after references.

@@

% Shorter version of title entered in capital letters:
\titlerunninghead{Brazilian Smoothing Seismicity and PSHA}

@@

% Corresponding author mailing address and e-mail address:
\authoraddr{Corresponding author: Marlon Pirchiner, Department of Geophysics, IAG, USP, São Paulo, Brazil.
(marlon.pirchiner@iag.usp.br)}

@@

Intraplate seismicity is a hard problem to solve in tectonics, 
main during the lack of observed seismicity in many areas where seismic 
hazard assesment are also difficult and the methods applied are often 
close correlated to the geometries defined by the experts with valuable 
criteria.

@@

All the methods are based on a kernel density estimation. The main 
difference between them is the \emph{bandwidth selection} process.

The Frankel method use a \emph{fixed} bandwidth in space. Woo method
suggest a bandwidth as function of the \emph{magnitude}.
Helmstetter and Werner proposes a \emph{local-adaptive} 
in space and time bandwidth, chosen for each event. 
It was proposed initialy to take care of the background seismicity rate 
on long-term earthquake forecast context, but the one year forecasted 
seismic rate could be a grid of points seismic sources under a 
time-independent assumption.

In the conclusions the results shown that

all methods gives good results and 
could be used in the Brazilian seismicity case,

all of them, 
% ###### improoving #########
the spatial resolution of the actual state-of-art GSHAP model.

The Last two methods was implemented and are available as a free software on HMTK from GEM Foundation. 
All the hazard calculations was performed on OpenQuake open source software.

@@

\subsection{Intraplate Seismicity}
\subsection{Previous Brazilian Studies}
\subsection{Previous Smoothing Seismicity (Frankel,Woo,Helmstetter\ldots)}
\subsection{Specifical Purpose}

@@

The Brazilain seismic hazard building code was made based on GSHAP model which is the well accepted state of art model
for the whole region.

@@

\section{Materials and Methods}

\subsection{Brazilian Seismic Catalog}
Boletim sismico brasileiro.

História e versão com Mw.
\subsection{Brazilian Seismic Catalog}

\begin{equation}
\label{eq:emc}
e = mc^2
\end{equation}

Vestibulum ante ipsum primis in faucibus orci luctus et ultrices posuere cubilia Curae; Mauris eu sapien nunc, sit amet accumsan dui. Nulla ac diam ut nunc placerat semper eget et libero. Vestibulum ante ipsum primis in faucibus orci luctus et ultrices posuere cubilia Curae; Cras hendrerit ullamcorper sapien vitae luctus. Quisque vel diam massa. Vestibulum dui nibh, facilisis vel vestibulum eu, viverra in quam.

\begin{eqnarray}
  x_{1} & = & (x - x_{0}) \cos \Theta \nonumber \\
        && + (y - y_{0}) \sin \Theta  \nonumber \\
  y_{1} & = & -(x - x_{0}) \sin \Theta \nonumber \\
        && + (y - y_{0}) \cos \Theta.
\end{eqnarray}

@@

Referencing equation \ref{eq:emc}. Mauris vel lorem magna, tristique auctor ipsum. Aliquam pharetra eleifend massa. Donec porttitor sagittis luctus. Aliquam pretium luctus leo quis congue. Morbi vel felis mi. Referencing Table \ref{sampletable}. Referencing Figure \ref{placeholder}.

@@

\begin{figure}
\includegraphics[width=0.4\linewidth]{placeholder.jpg}
\caption{Figure caption}\label{placeholder}
\end{figure}

\begin{table}
\caption{Table caption}\label{sampletable}
\begin{tabular}{l l l}
\hline
\textbf{Treatments} & \textbf{Response 1} & \textbf{Response 2}\\
\hline
Treatment 1 & 0.0003262 & 0.562 \\
Treatment 2 & 0.0015681 & 0.910 \\
Treatment 3 & 0.0009271 & 0.296 \\
\hline
\end{tabular}
\end{table}

@@

Math coded inside display math mode \[ ...\]
 will not be numbered, e.g.,:
 \[ x^2=y^2 + z^2\]

>>>>

%!TEX program = pdflatex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% American Geophysical Union (AGU)
% LaTeX Template
% Version 1.0 (3/6/13)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% The AGUTeX class and agu-ps referencing style were created and are owned
% by AGU: http://publications.agu.org/author-resource-center/author-guide/latex-formatting-toolkit/
%
% This template has been modified from the blank AGU template to include
% examples of how to insert content and drastically change commenting. The
% structural integrity is maintained as in the original blank template.
%
% Important notes:
% This template retains extensive commenting from the AGU template. It is heavily
% advised you read these comments and follow them in order to insure a speedy
% submission process.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% See how to enter figures and tables at the end of the article, after references.
\usepackage{float}
%\usepackage{caption}
\usepackage{subcaption}

%----------------------------------------------------------------------------------------
%	Glossaries
%----------------------------------------------------------------------------------------

% Glossaries
\usepackage[sanitize=none,acronym,toc]{glossaries}
%\usepackage[acronym,toc]{glossaries}
% Define a new glossary type
\newglossary[slg,toc]{symbols}{sym}{sbl}{Lista de Símbolos}
\newglossary[elg]{equations}{eqn}{eql}{Equations}
\makeglossaries

% ---------------------------------------------------------------------------- %
% Dicionario de Termos:
\input{d01-terms}
\input{d02-symbols}
\input{d03-acronyms}
\input{d04-equations}
% ---------------------------------------------------------------------------- %

% ---------------------------------------------------------------------------- %
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\erf}{erf}
% ---------------------------------------------------------------------------- %

@@

% Shorter version of title entered in capital letters:
\titlerunninghead{Smoothed seismicity Brazilian PSHA}

@@

% Corresponding author mailing address and e-mail address:
\authoraddr{Corresponding author: Marlon Pirchiner, Department of Geophysics, IAG, USP, S\~ao Paulo, Brazil.
(marlon.pirchiner@iag.usp.br)}

@@

Intraplate seismicity is a hard problem to solve in tectonics,
mainly during the lack of observed seismicity in many areas where seismic
hazard assesment are also difficult and the methods applied are often
close correlated to the geometries defined by the experts with valuable
criteria.

@@

All this methods are based on a kernel density estimation. The main
difference between them is the bandwidth selection process.

Frankel in their method uses a \emph{fixed} bandwidth in space. Woo's method
suggest a bandwidth as function of the \emph{magnitude}.
Helmstetter and Werner proposes a \emph{local-adaptive}
in space and time kernel bandwidths, chosen for each event.
This last one was proposed initialy to take care of the background seismicity rate
on long-term earthquake forecast context and the time-independent assumption
about the annual background seismic rate could be used
(not indiscriminately REFERENCE) as a grid of points seismic sources.

As conclusions the results shown that all methods gives defines better
some known seismic regions and could be used in the Brazilian seismicity case,
all of them, improoving the spatial resolution of the GSHAP model, the last one
made on the region.

The Last two methods was implemented and are available as a free software on
the Hazard Modeling Toolkit (HMTK) from GEM Foundation.

All the hazard calculations was performed on OpenQuake open source software.

@@

@@

The Brazilian seismic hazard building code was made based on GSHAP model which is the well accepted state of art model
for the whole region.

@@

\subsection{Intraplate Seismicity}

Onononono on on ono on on on on ononon on on on on on ono non on ono on on

\subsection{Previous Brazilian Studies}

Onononono on on ono on on on on ononon on on on on on ono non on ono on on

\subsection{Previous Smoothing Seismicity (Frankel, Woo, Helmstetter\ldots)}

Onononono on on ono on on on on ononon on on on on on ono non on ono on on

\subsection{Specifical Purpose}

%------------------------------------------------
\section{Data}

%------------------------------------------------
\section{Methods}

\subsection{Frankel}

A proposta de Arthur \citet{frankel_1995} foi usar uma largura de banda fixa,
nomeada distância de correlação $d_F$, e aplicar o estimador de Nadaraya-Watson (\ref{sec:nadaraya})
para suavizar o histograma 2D da sismicidade utilizando uma função de núcleo gaussiana:
\begin{equation}
	\ensuremath{
		\tilde{n}_j = \frac{ \sum_{i} n_i \,e^{ - \left(\frac{\gls{sym:dij}}{\gls{sym:dF}}\right)^2}}
						   { \sum_{i}     e^{ - \left(\frac{\gls{sym:dij}}{\gls{sym:dF}}\right)^2}},
	}
	\label{eq:ni}
\end{equation}
onde $\tilde{n}_j$ é a taxa de sismicidade (número de sismos com magnitude $m$ maior que a mínima magnitude
\gls{sym:Md} do catálogo) suavizada
na célula $j$, $n_i$ é o número de sismos em cada outra célula $i$ e
	\gls{sym:dij} é \glsdesc{sym:dij}.

\subsection{Woo}

Já Gordon \citet{woo_1996} propôs avaliar a contribuição de uma função de núcleo
de cada sismo $i$ ocorrido em $\boldsymbol{r}_i$
na célula centrada em $\boldsymbol{r}$ que dependa também de sua
magnitude $m$:
\begin{equation}
	\ensuremath{
		\gls{sym:Rrm} = \sum_{i=1}^{N} \frac{ K(\gls{sym:r} - \gls{sym:ri}, m)}
											{T({\gls{sym:ri})}},
	}
	\label{eq:Rrm}
\end{equation}
onde $N$ é o número de tremores $i$ no catálogo
e $T(\gls{sym:ri})$ é o período em que todo sismo de magnitude acima de $m$ é completamente observado
em \gls{sym:ri}.

\subsection{Helmstetter}

\begin{equation}
	\ensuremath{\gls{sym:R} = \sum_{i=1}^{N}{ \frac{1}{h_i\,{d_i}^2} \gls{sym:Kt}\gls{sym:Kr} }},
	\label{eq:helms01}
\end{equation}
onde \gls{sym:R} é \glsdesc{sym:R},
	  $K_t$ é a \glsdesc{sym:Kt},
	  $K_r$ é a \glsdesc{sym:Kr}.

Como seu interesse é em projeção, restringiram os tempos do catálogo aos $t_i
< t$ e ponderaram segundo um peso $w$ da seguinte maneira:
\begin{equation}
\ensuremath{\gls{sym:R} = \gls{sym:Rmin} + \sum_{t_i < t}{
	\frac{2\,w(\boldsymbol{r}_i,t_i)}{h_i\,{d_i}^2}
			\gls{sym:Kt}\gls{sym:Kr} }},
	\label{eq:helms02}
\end{equation}
onde \gls{sym:Rmin} é a \glsdesc{sym:Rmin}, positiva, permitindo que ocorram eventos-surpresa
onde não se tem registro de sismicidade.

Os pesos $w(\boldsymbol{r}_i,t_i)$, calculados em cada tremor $i$, propostos por Helmstetter são a transformação de
Gutemberg-Richter. Consideraram uma contribuição maior para a taxa de sismicidade daqueles sismos
que ocorreram onde e quando a magnitude de completude $M_c$ era maior que
a mínima magnitude $M_d$ do catálogo
em locais onde a magnitude de completude e o \emph{valor-b} variam.
Note que o modelo de pesos acomoda bem variações espaciais e
temporais tanto da magnitude de completude como do \emph{valor-b}. A expressão é a seguinte:
\begin{equation}
	\ensuremath{ w(\boldsymbol{r},t) = 10^{ \gls{sym:b}(\boldsymbol{r},t) \left[ \gls{sym:Mc_rt} - \gls{sym:Md}
	\right] } },
	\label{eq:helms_wi}
\end{equation}
onde  \gls{sym:wi} é o \glsdesc{sym:wi} na localização $\boldsymbol{r}$ e no instante $t$,
	  \gls{sym:b}$(\boldsymbol{r},t)$ é o \glsdesc{sym:b},
	  \gls{sym:Mc_rt} é a \glsdesc{sym:Mc_rt},
	  \gls{sym:Md} é a \glsdesc{sym:Md}.

\subsubsection{Bandwidth computation}

\begin{equation}
	\ensuremath{
%		h_i, d_i = \underset{d_i \ge \gls{sym:dk}, h_i \ge \gls{sym:hk}}{\argmin}
		h_i, d_i = \argmin_{\substack{h_i \ge \gls{sym:hk} \\
						              d_i \ge \gls{sym:dk}}
				           }
		\left[ s \left(h_i,d_i
			 		  \arrowvert
					  \gls{sym:k_cnn},\gls{sym:a_cnn}
			     \right)
			   := h_i + \gls{sym:a_cnn}d_i
	    \right],
	}
	\label{eq:helms_cnn}
\end{equation}
onde \gls{sym:k_cnn} é o \glsdesc{sym:k_cnn},
	 \gls{sym:a_cnn} é o \glsdesc{sym:a_cnn},
	 \gls{sym:dk} é o \glsdesc{sym:dk} e
	 \gls{sym:hk} é o \glsdesc{sym:hk}.

As larguras de bandas são escolhidas localmente, podendo ser menor, aumentando a resolução
onde há maior densidade de informação, ou crescer onde há escassez de informação e menor resolução.

\subsubsection{Taxa de sismicidade estacionária}
Mesmo havendo modelado a dependência com o tempo para distribuição da taxa de sismicidade,
a taxa estacionária $\bar{R}$ que se preserva no longo-prazo em uma determinada localização
$\boldsymbol{r}_0$ é a que mais interessa à \gls{psha} e,
segundo o método proposto por \citet{helmstetter_2012}, será definida
pelo valor da mediana da taxa fornecida pelo modelo no
período considerado:

\begin{equation}
	\ensuremath{
		\bar{R}(\boldsymbol{r}_0) = \text{Mediana}\left[R(\boldsymbol{r}_0, t)\right].
	}
	\label{eq:stationary_rate}
\end{equation}

O fato de se tomar a mediana faz com que variações na taxa de sismicidade devido a pré e pós-eventos
não interfiram significativamente no valor da taxa estacionária. A decorrência mais importante desse fato
é que o método dispensa a remoção de agrupamentos (seção \ref{sec:declustering}) do catálogo, um método paramétrico e
quase sempre controverso.

\subsubsection{Verossimilhança}

Para otimizar os parâmetros do modelo de Helmstetter, é preciso dividir o catálogo de sismos.
Uma parte serve para o aprendizado do modelo e a outra para avaliação.

Se a ocorrência de tremores pode ser modelada por um processo de Poisson com taxa $N_p$,
então a probabilidade de se observar exatamente $n$ eventos no período de tempo considerado é dada por

\begin{equation}
	\ensuremath{
		\gls{sym:pNn} = \frac{{N_p}^n e^{-N_p}}{n!}.
	}
	\label{eq:poisson_probability}
\end{equation}

E o logarítmo, a ser maximizado, da verossimilhança entre
o que o modelo predisse e o que foi observado
pode ser escrito como

\begin{equation}
	\ensuremath{
		\gls{sym:L} = \sum_{i_x=1}^{N_x}\sum_{i_y=1}^{N_y}\log p\left[  \gls{sym:Np}, \gls{sym:nxy}  \right]
	}
	\label{eq:loglik}
\end{equation}
onde \gls{sym:Np} é \glsdesc{sym:Np},
	\gls{sym:nxy} é \glsdesc{sym:nxy}.

Os parâmetros $R_{min}$, $a_{cnn}$ e $k_{cnn}$ do modelo são otimizados
pela maximização da verossimilhança $L$.

@@

Referencing equation \ref{eq:emc}. Mauris vel lorem magna, tristique auctor ipsum. Aliquam pharetra eleifend massa. Donec porttitor sagittis luctus. Aliquam pretium luctus leo quis congue. Morbi vel felis mi. Referencing Table \ref{fig_seismicity}. Referencing Figure \ref{placeholder}.

@@

% =========
% intro
% =========

\begin{figure}
	\centering
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=0.4\textwidth]{placeholder}
		\caption{caption A}
		\label{placeholder_a}
	\end{subfigure}
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=0.4\textwidth]{placeholder}
		\caption{caption B}
		\label{placeholder_b}
	\end{subfigure}
	\caption{main caption}
	\label{placeholder}
\end{figure}

\begin{figure}
	\includegraphics[width=0.4\linewidth]{z_img_seismicity_br}
	\caption{Brazilian earthquake catalog.}
	\label{fig_seismicity}
\end{figure}

% =========
% data
% =========

\begin{figure}
	\includegraphics[width=0.4\linewidth]{z_img_hmtk_bsb2013_rate}
	\caption{Earthquake records by year}
	\label{fig_seismic_rate}
\end{figure}

\begin{figure}
	\includegraphics[width=0.4\linewidth]{z_img_time_mag_count_br_1960}
	\caption{Magnitude count by year.}
	\label{fig_time_mag_count}
\end{figure}

\begin{figure}
	\includegraphics[width=0.4\linewidth]{z_img_decluster_br}
	\caption{Catalog declustering.}
	\label{fig_decluster}
\end{figure}

\begin{figure}
	\includegraphics[width=0.4\linewidth]{z_img_occurrence}
	\caption{Annual earthquake recurrence.}
	\label{fig_occurrence}
\end{figure}

% =========
% Methods
% =========

\begin{figure}
	\includegraphics[width=0.4\linewidth]{z_img_a_frankel_br}
	\caption{Smoothed seismicity: Frankel's method.}
	\label{fig_a_frankel}
\end{figure}

\begin{figure}
	\includegraphics[width=0.4\linewidth]{z_img_woo_bandwidth}
	\caption{Magnitude dependence bandwidth.}
	\label{fig_woo_bandwidth}
\end{figure}

\begin{figure}
	\includegraphics[width=0.4\linewidth]{z_img_a_woo}
	\caption{Smoothed seismicity: Woo's method.}
	\label{fig_a_woo}
\end{figure}

\begin{figure}
	\includegraphics[width=0.4\linewidth]{z_img_helmstetter_hidi}
	\caption{Local bandwidth example.}
	\label{fig_helmstetter_hidi}
\end{figure}

% TODO add HI_DI histograms

\begin{figure}
	\includegraphics[width=0.4\linewidth]{z_img_helmstetter_stationary_a}
	\caption{Stationary seismic rate}
	\label{fig_helmstetter_stationary_rate}
\end{figure}

\begin{figure}
	\includegraphics[width=0.4\linewidth]{z_img_helmstetter_catalogues}
	\caption{Learning and target catalogues.}
	\label{fig_helmstetter_catalogues}
\end{figure}

\begin{figure}
	\includegraphics[width=0.4\linewidth]{z_img_a_helmstetter}
	\caption{Smoothed seismicity: Helmstetter's method.}
	\label{fig_a_helmstetter}
\end{figure}

% =========
% Results
% =========

\begin{figure}
	\includegraphics[width=0.4\linewidth]{z_img_pga_gshap}
	\caption{GSHAP results: PGA (poe 10\%/50y)}
	\label{fig_pga_gshap}
\end{figure}

\begin{figure}
	\includegraphics[width=0.4\linewidth]{z_img_pga_frankel}
	\caption{Frankel's results: PGA (poe 10\%/50y)}
	\label{fig_pga_frankel}
\end{figure}

\begin{figure}
	\includegraphics[width=0.4\linewidth]{z_img_pga_woo}
	\caption{Woo's results: PGA (poe 10\%/50y)}
	\label{fig_pga_woo}
\end{figure}

\begin{figure}
	\includegraphics[width=0.4\linewidth]{z_img_pga_helmstetter}
	\caption{Helmstetter's results: PGA (poe 10\%/50y)}
	\label{fig_pga_helmstetter}
\end{figure}

% TABLES

\begin{table}[H]
	\caption{Table caption}
	\begin{tabular}{l l l}
		\hline
		\textbf{Treatments} & \textbf{Response 1} & \textbf{Response 2}\\
		\hline
		Treatment 1         & 0.0003262           & 0.562 \\
		Treatment 2         & 0.0015681           & 0.910 \\
		Treatment 3         & 0.0009271           & 0.296 \\
		\hline
	\end{tabular}
	\label{sampletable}
\end{table}

@@

Math coded inside display math mode \[ ...\]
 will not be numbered, e.g.,:
 \[ x^2=y^2 + z^2 \]
